Iterative Monomorphisation

### 2024-07-12 iterative_monomorphisation.tex
more numbers

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds, however, this is unfeasible. Instead, we group closely related bounds together and test combinations of values for bounds in these groups. Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected. Winners are shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.

As a precaution against overfitting, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems. We carried out the rest of the evaluation on the remaining 1034 problems.

@@

We conducted additional informal evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds. We started the option evaluation process with the base configuration and the following default values for the monomorphisation bounds
and the E invocation timeout:

\begin{multicols}{2}% 3 columns?
\begin{itemize}
   \item \PolyCap: \(\infty\)
   \item \PolyMult: \(1\)
   \item \PolyFloor: \(50\)
   \item substitution ordering: separation
   \item \SubstLimit: \(\infty\)
   \item \MonoSubstsLimit: \(10\)
   \item \Limit: \(2000\)
   \item new formulae limit multiplier: \(0\)
   \item monomorphisation timeout: \(20\)
   \item \Loop: \(4\)
   \item E timeout: \(30\)
   \item[] ~
\end{itemize}
\end{multicols}
\noindent
The initial values for the \MonoCap, \MonoMult and \MonoFloor options are
irrelevant because the options' values are set when computing
Table~\ref{mono_ty_args} and used for subsequent tables.

In all our experiments, monomorphisation and the subsequent call to E occur at the beginning of the invocation of Zipperposition. Departing from the original \verb|40_b.comb| configuration, which would otherwise call E after 15\% of the total alloted time for Zipperposition has passed. All others options either appear explicitely in the following tables or have been left unmodified.

@@

Table \ref{e_settings} evaluates the options with which we call the E prover. The call point corresponds to a fraction of the total time alloted to Zipperposition. Zipperposition is run for \(\text{timeout} \times \text{call point}\) seconds before monomorphisation and the subsequent call to E occur. The E timeout limits how lond E is run before being interrupted and resuming the normal operation of Zipperposition. The longer Zipperposition is run, the more formulae are generated, when taking into account the explosive nature of monomorphisation ...

\begin{table}[th]
\caption{Evaluation of parameters related to the E prover}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & \multicolumn{4}{c}{E call point} \\
   \multirow{1}{4em}{E timeout} & 0.0 & 0.1 & 0.2 & 0.3\\
   \midrule
   2      & 180 & 143 & 132 & 124 \\
   5   &\bf{185}& 142 & 134 & 125 \\
   10     & 184 & 143 & 132 & 125 \\
   20     & 184 & 137 & 133 & 125 \\
   30     & 182 & 133 & 134 & 125 \\
   \bottomrule
\end{tabular}
\label{e_settings}
\end{table}

@@

We compared the performance of two instances of Zipperposition. The first instance is run in the portfolio mode \verb|portfolio.sequential.py| which attempts to prove the problem with various configuration in succession. Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic. Because all given problems are non-monomorphic, this instance is labeled ``without E''. The second instance, is a modification of \verb|portfolio.sequential.py| where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase. This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately. The results of these evaluations seem to suggest that some amount of overfitting took place during the option optimisation phase because the number of problems by the version without E but not by the version with E is disproportionately higher in the 1034 problems benchmark relative to the 500 problems benchmark.

\begin{table}[ht]
\caption{Evaluation of Zipperposition without E vs. with E}
\centering\begin{tabular}{@{}lccc@{}}
   \toprule
   & without E & with E & Union \\
   \midrule
   500 problems benchmark   & 160 & 198 & 207 \\ % -9 +47
   fresh problems benchmark & 337 & 410 & 434 \\ % -24 +97
   \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[ht]
\caption{Evaluation of native polymorphism vs.\ monomorphisation}
\centering\begin{tabular}{@{}lccc@{}}
   \toprule
   & Native & Mono & Union \\
   \midrule
   E  &   & 0 & 0 \\
   Leo-III &  157 with E & 0 & 0 \\ 
   %Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
   Satallax &  & 0 & 0 \\
   %Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
   Zipperposition & 339 & 0 & 0 \\[1.5\jot]
   Total & 0 & 0 & 0 \\
   \bottomrule
\end{tabular}
\end{table}

>>>>

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds, however, this is unfeasible. Instead, we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected. Winners are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.

As a precaution against overfitting, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems. We carried out the rest of the evaluations on the remaining 1034 problems.

@@

We conducted additional informal evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds. We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:

\begin{multicols}{2}% 3 columns?
\begin{itemize}
   \item \PolyCap: \(\infty\)
   \item \PolyMult: \(1\)
   \item \PolyFloor: \(50\)
   \item substitution ordering: separation
   \item \SubstLimit: \(\infty\)
   \item \MonoSubstsLimit: \(10\)
   \item \Limit: \(2000\)
   \item new formulae limit multiplier: \(0\)
   \item monomorphisation timeout: \(20\)
   \item \Loop: \(4\)
   \item E timeout: \(30\)
   \item E call point: \(0\)
\end{itemize}
\end{multicols}
\noindent
The initial values for the \MonoCap, \MonoMult and \MonoFloor options are irrelevant because the options' values are set when computing Table~\ref{mono_ty_args}.% and used for subsequent tables.

@@

Table \ref{e_settings} evaluates the options with which we call the E prover. The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process. The E timeout limits how lond E is run before being interrupted and resuming the normal operation of Zipperposition. The longer Zipperposition is run, the more formulae are generated this and the explosive nature of iterative monomorphisation likely explains the poor performance of greater values of E call point.

\begin{table}[th]
\caption{Evaluation of parameters related to the E prover}
\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
   \toprule
   & \multicolumn{4}{c}{E call point} \\
   \multirow{1}{4em}{E timeout} & 0 & 0.1 & 0.2 & 0.3\\
   \midrule
   2      & 180 & 143 & 132 & 124 \\
   5   &\bf{185}& 142 & 134 & 125 \\
   10     & 184 & 143 & 132 & 125 \\
   20     & 184 & 137 & 133 & 125 \\
   30     & 182 & 133 & 134 & 125 \\
   \bottomrule
\end{tabular}
\label{e_settings}
\end{table}

@@

We compared the performance of two instances of Zipperposition. The first instance is run in the portfolio mode \verb|portfolio.sequential.py| which attempts to prove the problem with various configuration in succession. Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic. Because all given problems are non-monomorphic, this instance is labeled ``without E''. The second instance, is a modification of \verb|portfolio.sequential.py| where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase. This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately. The results of these evaluations seem to suggest that no significant overfitting took place during the option optimisation phase.

Table \ref{zipp_eval} shows that the use of E as a backend markedly improves the performance of Zipperposition. It is not a strict improvement as a number of problems are solved without E and not with E.

\begin{table}[ht]
\caption{Evaluation of Zipperposition without E vs. with E}
\centering\begin{tabular}{@{}lccc@{}}
   \toprule
   & without E & with E & Union \\
   \midrule
   500 problems benchmark   & 160 & 198 & 207 \\ % -9 +47
   fresh problems benchmark & 337 & 410 & 434 \\ % -24 +97
   \bottomrule
   \label{zipp_eval}
\end{tabular}
\end{table}

@@

\begin{table}[ht]
\caption{Evaluation of native polymorphism vs.\ monomorphisation}
\centering\begin{tabular}{@{}lccc@{}}
   \toprule
   & Native & Mono & Union \\
   \midrule
   E              &   & 340 & 340 \\
   Leo-III with E &  157 & 231 & 0 \\ 
   Zipperposition & 339 & 351 & 404 \\[1.5\jot]
   Total          & 0 & 0 & 0 \\
   % Satallax       &  & 0 & 0 \\ % doesn't understand negated conjectures
   %Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
   %Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
   \bottomrule
\end{tabular}
\end{table}
